---
title: "For Probabilistic Prediction, Full Bayes is Better than Point
Estimators"
author: "Bob Carpenter"
date: "August 2020"
output:
  tufte::tufte_html:
    toc: true
    toc_depth: 1
---

```{r echo = FALSE, warning = FALSE, message = FALSE}
library(ggplot2)
library(knitr)
library(kableExtra)
library(magrittr)
knitr::opts_chunk$set(
  include = TRUE,
  cache = TRUE,
  collapse = TRUE,
  echo = FALSE,
  message = FALSE,
  tidy = FALSE,
  warning = FALSE,
  comment = "  ",
  dev = "png",
  dev.args = list(bg = '#FFFFF8'),
  dpi = 300,
  fig.align = "center",
  fig.width = 7,
  fig.asp = 0.618,
  fig.show = "hold",
  out.width = "90%"
)
library(MASS)
library(rstan)
  rstan_options(auto_write = FALSE)
  options(mc.cores = 4)
# library(tufte)


# UTILITY FUNCTIONS

printf <- function(pattern, ...) {
  cat(sprintf(pattern, ...))
}

print_file <- function(file) {
  cat(paste(readLines(file), "\n", sep=""), sep="")
}

extract_one_draw <- function(stanfit, chain = 1, iter = 1) {
  x <- get_inits(stanfit, iter = iter)
  x[[chain]]
}

# TUFTE GGPLOT STYLE

ggtheme_tufte <- function() {
  theme(plot.background =
          element_rect(fill = "#fffff8",
                       colour = "#fffff8",
                       size = 0.5,
                       linetype = "solid"),
        plot.margin=unit(c(1, 1, 0.5, 0.5), "lines"),
        panel.background =
          element_rect(fill = "#fffff8",
                       colour = "#fffff8",
                       size = 0.5,
                       linetype = "solid"),
        panel.grid.major = element_line(colour = "white",
                                        size = 1, linetype="dashed"),
        panel.grid.minor = element_blank(),
        legend.box.background =
          element_rect(fill = "#fffff8",
                       colour = "#fffff8",
                       linetype = "solid"),
        axis.ticks = element_blank(),
        axis.text = element_text(family = "Palatino", size = 14),
        axis.title.x = element_text(family = "Palatino", size = 16,
                                    margin = margin(t = 15,
                                                    r = 0, b = 0, l = 0)),
        axis.title.y = element_text(family = "Palatino", size = 16,
                                    margin = margin(t = 0,
                                                    r = 15, b = 0, l = 0)),
        strip.background = element_rect(fill = "#fffff8",
                                        colour = "#fffff8",
                                        linetype = "solid"),
        strip.text = element_text(family = "Palatino", size = 14),
        legend.text = element_text(family = "Palatino", size = 14),
        legend.title = element_text(family = "Palatino", size = 16,
                                    margin = margin(b = 5)),
        legend.background = element_rect(fill = "#fffff8",
                                        colour = "#fffff8",
                                        linetype = "solid"),
        legend.key = element_rect(fill = "#fffff8",
                                        colour = "#fffff8",
                                        linetype = "solid")
  )
}

# GENERAL R CONFIGURATION

options(digits = 2)
options(htmltools.dir.version = FALSE)

set.seed(1234)  # reproduces with other seed, just tired of it dancing
```

## Abstract {-}

A probabilistic prediction takes the form of a distribution over possible outcomes.  With proper scoring rules such as log loss or square error, it is possible to evaluate such a probabilistic prediction against a true outcome.  This short note provides simulation-based evaluation of full Bayesian inference, where we average over our estimation uncertianty, and two forms of point estimation, one that uses the posterior mode (max a posteriori) and one that uses the posterior mean (as is typical with variational inference).  The example we consider is a simple Bayesian logistic regression with potentially correlated predictors and weakly informative priors.  To make a long story short, full Bayes has lower expected log loss and squared error than either of the point estimators.


# Logistic Regression

## Observed data

The data consists of binary observations $y_n \in \{ 0, 1 \}$ paired
with $K$-dimensional vectors of predictors^[Predictors are also called
covariates or features.] $x_n$ for $n \in 1:N$. That is, $x \in
\mathbb{R}^{N \times K}$ is the complete data matrix and $x_n$ is its
$n$-th row. We will assume the columns of $x$ are standardized,^[To
standardize an unstandardized variable $x = x_1, \ldots, x_N,$ set
$$\mbox{z}(x) = \frac{x - \mbox{mean}(x)}{\mbox{sd}(x)}.$$ A
standardized value $z(x)$ of 1 corresponds to an original value $x$
that is one standard deviation above the mean; a value of -2.5
corresponds to a value two and a half standard deviations below the
mean.] so that for every column $k$ corresponding to a predictor, we
have
$$
\mbox{mean}(x_{1:N, \ k}) = 0
$$
and
$$ \mbox{sd}(x_{1:N, \ k}) = 1.
$$


## Data distribution

For the purposes of simulation, we will assume the data have a
multivariate normal distribution with a positive-definite covariance
matrix $\Sigma$,
$$
x_n \sim \mbox{multinormal}(0, \Sigma).
$$
We will assume the covariance matrix $\Sigma$ has a unit diagonal so
that each $x_{n, \ k}$ has a standard normal marginal distribution.^[In
symbols, if $\mu = 0$ and $\mbox{diag}(\Sigma) = 1,$ then $y \sim
\mbox{multinormal}(\mu, \Sigma)$ implies that marginally, each $y_k
\sim \mbox{normal}(0, 1).$]

## Sampling distribution

The model is parameterized with an intercept $\alpha \in \mathbb{R}$
and coefficient vector $\beta \in \mathbb{R}^K$. Logistic regression
is a generalized linear model where the linear predictor

$$
\alpha + x_n \, \beta
\ = \
\alpha + \sum_{k=1}^K x_{n, k} \cdot \beta_k,
$$
represents the log odds of $y_n$ being equal to one.^[The function
$\mbox{logit}:(0, 1) \rightarrow (-\infty, \infty)$
maps a probability $v \in [0, 1]$ to its log odds,
$$\mbox{logit}(v) = \log \frac{v}{1 - v}.$$]

Given the log odds, the probability that $y_n$ is one is given by
inverting the log odds function,^[The inverse logit function
$\mbox{logit}^{-1}:(-\infty, \infty) \rightarrow (0, 1)$ maps a log
odds value $u \in (-\infty, \infty)$ to its corresponding probability,
$$\mbox{logit}^{-1}(u) = \frac{1}{1 + \exp(-u)}. The inverse logit
function is also known as the (logistic) sigmoid function.$$]
$$
\mbox{Pr}[y_n = 1] = \mbox{logit}^{-1}(\alpha + x_n \, \beta).
$$
The sampling distribution is then defined to follow the log odds,
$$
y_n \sim \mbox{bernoulli}(\mbox{logit}^{-1}(\alpha + x_n \, \beta)),
$$
for observations indexed by $n \in 1:N$. It is important to note that
the $y_n$ are defined by sampling according to the log odds.^[Because
it is used as a decision process to choose a response, a common
simulation mistake is to treat the log odds as a deterministic
threshold and define $$y_n = \begin{cases}1 & \mbox{if} \
\mbox{logit}^{-1}(\alpha + x_n \, \beta) > \frac{1}{2} \\[2pt] 0 &
\mbox{otherwise}.\end{cases}$$  Such an approach typically leads to
separable data leading to infinite maximum likelihood coefficients.
As some indication of how common this mistake is in simulating
logistic regression data, it
comes up several times in the first page of hits for the Google query (simulate data
with logistic regression), e.g.,
[(1)](https://stats.stackexchange.com/questions/282804/is-it-possible-to-simulate-logistic-regression-without-randomness)
[(2)](https://stats.stackexchange.com/questions/46523/how-to-simulate-artificial-data-for-logistic-regression/46525),
[(3)](https://www.statalist.org/forums/forum/general-stata-discussion/general/1475130-simulating-data-for-logistic-regression-with-categorical-variables?p=1475211#post1475211).]


## Prior distribution

Given the logistic scale and standardized predictors, we assume weakly
informative priors,^[Weakly informative priors provide the scale of a
parameter; for logistic regression, log odds outside of the -5 to 5
range have very low or very high probabilities.   For example,
$$\mbox{logit}^{-1}(6) \approx 0.9975.$$  So we keep the coefficients
in the -3 to 3 range to match the predictors in the same range so that
their products tend to not be too extreme on the log odds scale.]
$$
\alpha, \beta_k \sim \mbox{normal}(0, 2),
$$
for $k \in 1:K$.


## Predictive distribution

Suppose we also have some observed test predictors $\tilde{x}_n$ for
$n \in 1:\tilde{N}$, but we do not know the corresponding outcomes,
$\tilde{y}_n$. These outcomes need to be predicted based on the
predictors $\tilde{x}_n$ and the knowledge of the values of the
regression coefficients, $\alpha$ and $\beta$. Although not strictly
necessary, we will assume for the sake of simulation that
$\tilde{x}_n$ has the same distribution as the training predictors
$x$.^[If the conditions are not matched, poststratification can be
used to make collective predictions, such as estimating accuracy, on
mismatched data sets.]

If we knew the parameter values $(\alpha, \beta)$, we
would know that the distribution of $\tilde{y}_n$ is just the sampling
distribution,
$$
\tilde{y}_n \sim
\mbox{bernoulli}(\mbox{logit}^{-1}(\alpha + \tilde{x}_n \, \beta)).
$$


# Inference

In general, we do not know the regression coefficients $\alpha$ and
$\beta$;  we only observe data $(x, y)$ that lets us infer their
values with some uncertainty.  Similarly, we do not know the outcomes
$\tilde{y}_n$ we wish to predict based on their predictors
$\tilde{x}_n$ and the previously observed data $(x, y).$


## Bayesian inference

Full Bayesian inference averages^[For continuous quantities,
averages of functions $f(u)$ weighted by a probability function $p(u)$
for a random variable $U$ are given by expectations $$\mathbb{E}[f(U)]
\ = \ \int_U f(u) \cdot p(u) \, \mbox{d}u.$$]  over
the uncertainty in our parameters given the data.  Symbolically, this
can be expressed in terms of a posterior distribution,
as given by Bayes's rule,
$$
\begin{array}{rcl}
p(\alpha, \beta \mid x, y)
& = &
\frac{\displaystyle p(y \mid \alpha, \beta, x) \cdot p(\alpha, \beta)}
     {\displaystyle p(y)}
\\[4pt]
& \propto &
p(y \mid \alpha, \beta, x) \cdot p(\alpha, \beta).
\end{array}
$$
with terms for the sampling distribution $p(y \mid \alpha, \beta, x)$
and prior $p(\alpha, \beta).$

What we really want is inference for $\tilde{y}$, the unobserved
outcomes. These are defined by taking a weighted average of
predictions $p(\tilde{y} \mid \tilde{x}, \alpha, \beta)$, where the weights are determined by the posterior
density $p(\alpha, \beta \mid x, y)$,
$$
p(\tilde{y} \mid \tilde{x}, x, y)
\ = \
\int_{\mathbb{R}} \int_{\mathbb{R}^K}
\,
p(\tilde{y} \mid \tilde{x}, \alpha, \beta)
\cdot
p(\alpha, \beta \mid x, y)
\,
\mathrm{d} \beta
\,
\mathrm{d} \alpha.
$$

Given posterior draws $\left( \alpha^{(m)}, \beta^{(m)} \right)$,^[These may be
(anti-)correlated, as in the draws produced by Markov chain Monte
Carlo.] the posterior predictive distribution may be calculated by
plugging in draws and averaging,
$$
p(\tilde{y} \mid \tilde{x}, x, y)
\ \approx \
\frac{1}{M} \sum_{m = 1}^M p(\tilde{y} \mid \tilde{x}, \alpha^{(m)}, \beta^{(m)}).
$$
Expected error in the estimate decreases as $\mathcal{O}(1 /
\sqrt{M}).$^[Technically, the denominator the square root of the
effective sample size, but this is linearly related to $M$, so the
result holds as stated up to an order.]



## Max a posterior approximate inference

The maximum a posteriori (MAP) estimator for parameters $(\alpha,
\beta)$ is given by
$$
\alpha^*, \beta^*
\ = \
\mbox{arg max}_{\alpha, \ \beta}
\
p(\alpha, \beta \mid x, y).
$$

It is common in machine learning applications to simply plug-in
$(\alpha^*, \beta^*)$ for inference, using the approximation
$$
p(\tilde{y} \mid \tilde{x}, x, y)
\ \approx \
p(\tilde{y} \mid \tilde{x}, \alpha^*, \beta^*).
$$

The approximation arises because the values $\alpha^*, \beta^*$ are
being treated as certain when the data $(x, y)$ provide only limited
information about their values.


## Variational approximation inference

Variational Bayes (VB) attempts to find an approximate distribution
matching the posterior. In machine learning settings, the goal is
typically to extract point estimates corresponding to the approximate
posterior mean values.^[It's possible to extract posterior simulation
draws given the full variational approximation, but that is almost
never done other than in Stan. Typically, variational approximations
are mean field (diagonal covariance), and so the posterior
approximation is poor when parameters are correlated, as in typical
logistic regression applications to language, vision, survey questions,
or other correlated system of predictors. It is typical instead to
just plug in the approximate posterior means found by variational
inference.] If the VB approximation were perfect, the optimizer will
find the true posterior means.^[This doesn't matter if it uses a
mean-field approximation or not; best case, it recovers the true
posterior means.] For the purposes of simulation, we are going to
suppose we have a good enough variational approximation to find the
true posterior means.^[This is very unlikely in practice given the
nature of the variational approximations, but given the
variety of variational approximations possible and algorithms to fit
them, this seemed like a good compromise for evaluation.]

We can define the posterior means directly as conditional posterior
expectations,
$$
\begin{array}{rcl}
\widehat{\alpha}, \widehat{\beta}
& = & \mathbb{E}[\alpha, \beta \mid x, y]
\\[8pt]
& = & \int_{\mathbb{R}} \int_{\mathbb{R}^K} (\alpha, \beta) \cdot
p(\alpha, \beta \mid x, y) \, \mathrm{d} \beta \, \mathrm{d} \alpha.
\end{array}
$$
We can calculate the exact posterior means to arbitrary precision
using simulated draws $\alpha^{(m)}, \beta^{(m)}$ from the
posterior,
$$
\begin{array}{rcl}
\widehat{\alpha} & = & \displaystyle \frac{1}{M} \sum_{m = 1}^M \alpha^{(m)}
\\[12pt]
\widehat{\beta} & = & \displaystyle \frac{1}{M} \sum_{m = 1}^M \beta^{(m)}.
\end{array}
$$

Unlike maximum likelihood estimates, posterior mean estimates are
unbiased and have the pleasant property of minimizing expected square
error in the parameter estimates (we define square error below).

Given the posterior means, we can define another point-based
approximation of the predictive distribution,
$$
p(\tilde{y} \mid \tilde{x}, x, y)
\ \approx \
p(\tilde{y} \mid \tilde{x}, \widehat{\alpha}, \widehat{\beta}).
$$

As with posterior mode estimates, posterior mean estimates are
approximate in that they assume that $\widehat{\alpha},
\widehat{\beta}$ are estimated correctly, whereas there is residual
uncertainty after observing the training data $(x, y).$

# Simulation Experiment

In this section, we're going to simulate some correlated predictors
and outcomes and then fit our logistic regression model with full
Bayes, posterior modes, and posterior means for the purposes of
evaluating predictive accuracy.

## Stan program

Here's a Stan program implementing the logistic regression model.
```{r}
print_file('logistic-regression.stan')
```
The program includes not only the data and model declaration, but also
the predictive distributions for square error and log loss. The true
values of the test cases are provided to the program, but they are not
used during training.^[Stan's generated quantities block is executed
each iteration based on the parameter draws for that iteration; no
information in the generated quantities block flows back to parameter
estimation. For semi-supervised learning in cases such as naive Bayes
or HMMs where there isn't the factorization of predictors found in
logistic regression, we would need to move the tests into the model
block to provide full Bayesian inference that also uses information in
the test predictors (but not the test outcomes).]


## Simulated Data

We'll assume the predictor vectors are multivariate normal with
$K \times K$ covariance matrix $\Sigma$ defined by
$$
\Sigma_{i, j} = \rho^{|i - j|},
$$
for some $\rho \in (0, 1).$

```{r}
K <- 50
rho <- 0.9
Sigma <- matrix(0, K, K)
for (i in 1:K) {
  for (j in 1:K) {
    Sigma[i, j] <- rho^abs(i - j)
  }
}
```

```{r}
inv_logit <- function(u) 1 / (1 + exp(-u))

alpha <- rnorm(1, 0, 0.5)
beta <- rnorm(K, 0, 0.5)

N <- 200
N_test <- 500
x <- mvrnorm(N, rep(0, K), Sigma)

y <- rbinom(N, 1, inv_logit(alpha + x %*% beta))

x_test <- mvrnorm(N_test, rep(0, K), Sigma)
y_test <- rbinom(N_test, 1, inv_logit(alpha + x_test %*% beta))

data <- list(K = K, N = N, x = x, y = y,
             N_test = N_test, x_test = x_test, y_test = y_test)
```
Specifically, we'll take $K = `r K`$, $\rho = `r rho`$, and $N = `r
N`$. Here's what the matrix looks like for $K = 5$ and $\rho = 0.9$.
$$
\Sigma \ = \
\left[
\begin{array}{ccccc}
1 & 0.9 & 0.81 & 0.73 & 0.66 \\
0.9 & 1 & 0.9 & 0.81 & 0.73 \\
0.81 & 0.9 & 1 & 0.9 & 0.81 \\
0.73 & 0.81 & 0.9 & 1 & 0.9 \\
0.66 & 0.73 & 0.81 & 0.9 & 1
\end{array}
\right]
$$

With highly correlated predictors, a predictor vector of size $K$
provides less information than if the predictors were not correlated.
How much less information depends on the degree of correlation---if
two predictors are perfectly correlated or anti-correlated, there is no new
information.

# Loss Functions for Evaluating predictions

Our goal is to provide predictive estimates of the probability that an
unobserved outcome $\tilde{y}_n$ takes value 1, given predictors
$\tilde{x}_n$ and training data $(x, y)$.  In symbols, we want to
estimate
$$
\mbox{Pr}\left[\tilde{y}_n = 1 \mid \tilde{x}_n, x, y\right].
$$
To do so, we want to use a *proper scoring rule*.^[Proper scoring
rules are ones which are optimized by the true distribution. Accuracy
(aka 0/1 loss), as is commonly used in machine learning, is not a
proper scoring rule.]  We will consider three proper scoring functions
(log loss, square loss, spherical loss) and one improper one (absolute
loss).

## Log loss

The simplest proper scoring rule is just the log probability (density
or mass) of the true result under the model.  For a given target $y_n$
and probabilistic prediction $\hat{y}_n$, the log loss is defined by

$$
\begin{array}{rcl}
\mbox{logloss}(y_n, \hat{y}_n)
& = & - \log \mbox{bernoulli}(y_n \mid \hat{y}_n)
\\[4pt]
& = & (y_n = 1) \ ? \ -\log(\hat{y}_n) \ : \ -\log(1 - \hat{y}_n).
\\[4pt]
& = & - \log (1 - \left| y_n - \hat{y}_n \right|).
\end{array}
$$
Extending to a complete test set $y$ and response set $\tilde{y}$,
$$
\mbox{logloss}(y, \tilde{y})
\ = \
\sum_{n=1}^{N} \mbox{logloss}(y_n, \hat{y}_n).
$$

We will report log loss as a rate per item after dividing by the
number of test items, $\tilde{N}.$

If we think of our test set elements $\tilde{y}_n$ as draws from the true
distribution $p(\tilde{y}_n \mid \tilde{x}_n)$, then log loss provides a
Monte Carlo estimate of the cross entropy rate from the true distribution of
outcomes to the probabilistic forecasts.^[Cross-entropy is defined
from a density $p$ to a density $q$ as
$$
\begin{array}{rcl}
\mathrm{H}[p, q]
& = &
- \int_Y p(y \mid x) \log q(y \mid x) \, \mathrm{d}y,
\\[8pt]
& \approx &
-\frac{1}{\tilde{N}} \sum_{n=1}^{\tilde{N}} \log q(\tilde{y}_n \mid \tilde{x}_n).
\end{array}
$$
]


## Square error

When the estimate takes a point form, as in our estimate of a
probability, we can compute square loss,^[Also known as a Brier score
in machine learning after Glenn Brier, the statistician who introduced
it in 1950.]
$$
\mbox{sqloss}(y_n, \tilde{y}_n)
\ = \
(y_n - \hat{y}_n)^2.
$$
Extending to an entire test set, the loss may be expressed compactly as
$$
\begin{array}{rcl}
\mbox{sqloss}(y, \hat{y}) =
& = & (y - \hat{y})^{\top} \, (y - \hat{y})
\\[4pt]
& = & \sum_{n=1}^{\tilde{N}} (y_n - \hat{y}_n)^2.
\end{array}
$$

Square loss is important in probability theory as it is the loss under
which the average is the optimal estimator.^[In symbols, for $x \in \mathbb{R}^N$,
$$\mbox{mean}(x) = \mbox{argmin}_{u} \sum_{n=1}^N (u - x_n)^2$$.]  Put
another way, the posterior mean is the point estimate that minimizes
expected square loss.

Typically, when considering square error, we consider the square error
rate by dividing by the number of test items, and then take the square
root of that to put it back on the natural probability scale; the
result is called "root mean square error" (RMSE),^[Taking means and
square roots are monotonic operations and do not change the ordering
of comparisons.]
$$
\mbox{rmse}(y, \hat{y})
\ = \
\sqrt{\frac{sqloss(y, \hat{y})}
           {\mbox{size}(y)}}
$$

## Spherical loss

Given an outcome $y_n$ and predicted probability $\hat{y}_n$, the
spherical loss is
$$
\mbox{sphereloss}(y_n, \hat{y}_n)
\ = \
1 - \frac{\displaystyle \left| y - \hat{y} \right|}
         {\displaystyle \sqrt{\hat{y}^2 + (1 - \hat{y})^2}}.
$$
Extending to a whole data set, we just sum the spherical losses,
$$
\mbox{sphereloss}(y, \hat{y})
\ = \
\sum_{n=1}^N \mbox{sphereloss}(y_n, \hat{y}_n).
$$

Although this is a proper loss function, we will not be evaluating
spherical loss because of the difficulty in interpretation---it is on
the natural scale of $y$, but adjusted by the skew in the prediction.
This reduces losses when predictive probabilities $\hat{y}$ are near
one half and increases losses when predictions are extreme (near zero
or one).


## Absolute error

The final loss function we consider is improper, and corresponds to
the absoute error, rather than square error.  The absolute value
function keeps the error positive,

$$
\mbox{absloss}(y_n, \tilde{y}_n)
\ = \
\left| \tilde{y}_n - \hat{y}_n \right|.
$$

For multiple $y_n$, we sum the absolute losses.  But we report them as
an average given that the values are on the natural probability scale.

Absolute error is *not* a proper probabilistic scoring rule.


## Plotting the loss functions

The absolute error can range between 0 and 1.  The square error has
the same range.  Log loss is minimized at zero, but unbounded as error
approaches one.  We can plot these functions

```{r fig.cap = "Three loss functions plotted as a function of absolute error.  Because the dimensions of the loss functions vary and the scale is irrelevant for comparison, we have multiplied the loss functions so that they have loss one half for an absolute error of one half.  For absolute errors between zero and one half, square error grows most slowly and absolute error increases the fastest.  For errors above one half, square error grows fastest initially, then log loss takes over for very high absolute error (above ninety percent or so)."}

err = seq(0.001, 0.999, by = 0.01)
err_df <-
  rbind(data.frame(error = err,  loss = err^2 * 0.5 / 0.5^2,
                   type = "sq error"),
        data.frame(error = err, loss = abs(err),
                   type = "abs error"),
        data.frame(error = err, loss = -log(1 - err) * (0.5 / -log(0.5)),
                   type = "log loss"))
err_plot <-
  ggplot(err_df, aes(x = error, y = loss, group = type)) +
  geom_line(aes(color = type)) +
  xlab("absolute prediction error") +
  ylab("normalized loss") +
  ggtheme_tufte()
err_plot
```
Each loss function has been multiplicatively normalized to provide a
loss of one half for an absolute error of one half while leaving a
loss of zero for an absolute error of zero.

It's also useful to observe the plot on the log scale to visualize the
behavior of the loss functions at values near zero.

```{r fig.cap = "Log scale version of the previous plot of loss versus absolute error.  The log scale highlights the differing behavior for small absolute errors."}
log_err_plot <-
  ggplot(err_df, aes(x = error, y = loss, group = type)) +
  geom_line(aes(color = type)) +
  scale_y_log10(breaks = c(0.000001, 0.001, 1),
                labels = c("0.000001", "0.001", "1")) +
  scale_x_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1),
                     labels = c(1, 0.75, 0.5, 0.25, 0)) +
  xlab("absolute prediction error") +
  ylab("normalized loss") +
  ggtheme_tufte()
log_err_plot
```


# Evaluation with Stan

Fitting the model we provided earlier to our simulate data in Stan
provides the following output, where we have only included the first
coefficient, $\beta_1$.
```{r results = 'hide'}
model <- stan_model("logistic-regression.stan")
fit <- sampling(model, data = data, iter = 4000,
                control = list(adapt_delta = 0.99, stepsize = 0.01),
                refresh = 0, open_progress = FALSE)
```
```{r}
print(fit, prob = c(0.01, 0.99),
           pars = c("alpha", "beta[1]", "log_loss", "sq_loss"))
```

The 1% and 99% values provide the ends of the central 98% posterior
interval on log loss and square loss for parameter values
$\alpha^{(m)}, \beta^{(m)}$, drawn from the posterior. With only $`r
N`$ training examples with correlated parameters, there is a great
deal of uncertainty in the posteriors for coefficients and hence for
predictions and for loss versus the test set. The range of log loss
and square loss values shows the variability in those metrics across
draws from the posterior.


## Full Bayesian inference

We first consider full Bayesian inference, where our estimate is
derived from
$$
p(\tilde{y}_n \mid \tilde{x}_n, x, y)
\ = \
\int p(\tilde{y}_n \mid \tilde{x}_n, \alpha, \beta)
\cdot p(\alpha, \beta \mid x, y) \, \mbox{d}(\alpha, \beta).
$$
The result of running the simulation yields the following losses.

```{r}
y_test_hat_bayes <- rep(0, N_test)
for (n in 1:N_test)
  y_test_hat_bayes[n] <- mean(extract(fit)$E_y_test[ , n])

sq_error_bayes <- sum((y_test - y_test_hat_bayes)^2)
log_loss_bayes <- -sum(dbinom(y_test, 1, y_test_hat_bayes, log = TRUE))

printf("BAYES: sq error %5.2f  root mean sq error %3.2f",
       sq_error_bayes, sqrt(sq_error_bayes / N_test))
printf("BAYES: log loss %5.2f  log loss rate %3.2f",
       log_loss_bayes, log_loss_bayes / N_test)
```

The root mean square error being reported is the square root of the
average square loss per test item. The log loss rate is just the log
loss divided by the number of test items.

## Variational inference

Using the draws from the posterior, we calculate the posterior means
$(\widehat{\alpha}, \widehat{\beta})$ and then plug those in to use
the estimate
$$
p(\tilde{y}_n \mid \tilde{x_n}, \widehat{\alpha}, \widehat{\beta}).
$$
The resulting errors are as follows.
```{r}
alpha_hat <- mean(extract(fit)$alpha)
beta_hat <- rep(0, K)
for (k in 1:K)
  beta_hat[k] <- mean(extract(fit)$beta[ , k])
y_test_hat_vi <- inv_logit(alpha_hat + (x_test %*% beta_hat))

sq_error_vi <- sum((y_test - y_test_hat_vi)^2)
log_loss_vi <- -sum(dbinom(y_test, 1, y_test_hat_vi, log = TRUE))

printf("VB: sq error %5.2f  root mean sq error %3.2f",
       sq_error_vi, sqrt(sq_error_vi / N_test))
printf("VB: log loss %5.2f  log loss rate %3.2f",
       log_loss_vi, log_loss_vi / N_test)
```

## Max a posterior inference

Here, we use Stan's optimizing function to find the maximum a
posteriori (MAP) estimates $\alpha^*, \beta^*$ and proceed to plug
those in for inference,
$$
p(\tilde{y}_n \mid \tilde{x_n}, \alpha^*, \beta^*).
$$
The resulting errors are as follows.

```{r}
fit_map <- optimizing(model, data = data, verbose = FALSE, refresh = 0)
alpha_hat_map <- fit_map$par[['alpha']]
beta_hat_map <- rep(0, K)
for (k in 1:K)
  beta_hat_map[k] <- fit_map$par[[paste('beta[', k, ']', sep = "")]]

y_test_hat_map <- inv_logit(alpha_hat_map + (x_test %*% beta_hat_map))

sq_error_map <- sum((y_test - y_test_hat_map)^2)
log_loss_map <- -sum(dbinom(y_test, 1, y_test_hat_map, log = TRUE))


printf("MAP: sq error %5.2f  root mean sq error %3.2f",
       sq_error_map, sqrt(sq_error_map / N_test))
printf("MAP: log loss %5.2f  log loss rate %3.2f",
       log_loss_map, log_loss_map / N_test)
```


# Probabilistic Training and Test Data

Traditionally, we estimate (i.e., train) logistic regression models
based on dichotomous observations
$$
y_n \in \{ 0, 1 \}.
$$

Now suppose instead that the outcomes $y_n$ are not known exactly, but
only with some uncertainty.  That is, rather than knowing $y_n$, we
only have a probability estimate of whether $y_n = 1$ or $y_n = 0.$
Such uncertainty can arise from noisy measurements of the true $y_n$
value.  For example, the data may be crowdsourced human data
coding,^[Also known as data rating, annotation, and labeling.] or it
may arise from the application of imprecise heuristics.
Alternatively, we may have uncertainty propagated from measurement
devices such as RNA-seq readers of genomic sequences.

The math all generalizes if we assume that instead of dichotomous
outcomes $y_n \in \{ 0, 1 \},$ we have probabilistic outcomes,
$$
\phi_n = \mbox{Pr}[y_n = 1 \mid x_n].
$$

Typically these $\phi_n$ are the estimates from another probabilistic
model, such as a noisy measurement model of human data coding or a
model of fluorescence response and image processing steps for
RNA-seq data.^[Ideally, the predictive variables $\phi$ estimated by the
training data measurement error model and the parameters $\alpha,
\beta$ of a regression based on that model will be modeled and
estimated jointly.]

## Weighted regression

Training is simple with weighted observations.  The log likelihood function for
dichotomous data $y_n \in \{ 0, 1 \}$ is as defined previously,

$$
\begin{array}{rcl}
\log p(y \mid x, \alpha, \beta)
& = &
\log \prod_{n = 1}^N p(y_n \mid x_n, \alpha, \beta)
\\[6pt]
& = &
\sum_{n = 1}^N
  \log
  \mbox{bernoulli}(y_n \mid \mbox{logit}^{-1}(\alpha + x_n \, \beta)).
\end{array}
$$

To generalize to continuous outcomes $\phi_n$, we work in expectation
with respect to the $\phi_n$.  So instead of the likelihood, we use a
weighted version,
$$
\sum_{n=1}^N
\begin{array}{l}
\phi_n
\cdot
\log \mbox{bernoulli}(1 \mid \mbox{logit}^{-1}(\alpha + x_n \, \beta))
\\
{ } + (1 - \phi_n)
\cdot
\log \mbox{bernoulli}(0 \mid \mbox{logit}^{-1}(\alpha + x_n \, \beta))
.
\end{array}
$$
This quantity simply replaces the likelihood in the model.  If the
values of $\phi_n$ are 0 or 1, it reduces to our original likelihood
function.


In Stan, our original vectorized sampling statement,
```
y ~ bernoulli_logit(alpha + x * beta);
```
is equivalent to the unrolled version
```
for (n in 1:N)
  target += bernoulli_logit(y[n] | alpha + x[n] * beta);
```
The weighted version computes the density as the expected log
likelihood given probability estimates $\phi_n = \mbox{Pr}[y_n = 1
\mid x_n],$
```
for (n in 1:N) {
  target += phi[n] * bernoulli_logit(1 | alpha + x[n] * beta);
  target += (1 - phi[n]) * bernoulli_logit(0 | alpha + x[n] * beta);
}
```
It's important that both target increment statements are used.  The
effect of incrementing both 0 and 1 results in proportion to their
weights provides a form of data-driven regularization, where the
resulting density is maximized when

$$
\phi_n = \mbox{logit}^{-1}(\alpha + x_n \, \beta).
$$

That is, we penalize coefficients for predicting probabilities for
$y_n$ that are too far away from the training set probabilities
$\phi_n$.  If we had sampled rather than weighted, the result would be
the same in the limit of increasing amounts of data.

The resulting formulation is not properly generative---there is no way
to generate the "observations" $\phi_n$ from the regression
coefficients $\alpha, \beta$ and the observed predictors $x_n.$

In this particular case, because we know that the log odds are given
by $\mbox{logit}(\phi_n)$, instead of using a weighted logistic regression,
we can simply train a standard linear regression
$$
\mbox{logit}(\phi_n) \sim \mbox{normal}(\alpha + x_n \, \beta, \sigma)
$$
and use the coefficients $\alpha, \beta$ predictively for new data
with a logit link function,
$$
\tilde{y}_n \sim \mbox{bernoulli}(\mbox{logit}^{-1}(\alpha + \tilde{x}_n \, \beta)).
$$

## Weighted logistic regression with Stan

Here's the full Stan model for weighted logistic regression.  It
follows the description in the previous section.  The data has been
modified to read in a vector of probabilities `phi[1:N]` rather than a
binary array `y[1:N].`
```{r}
print_file('weighted-logistic-regression.stan')
```

What we want to compare is how well we can estimate the model with
draws from the distribution given by $\phi$ versus estimating it using
the $\phi$ directly.  We will again simulate $N = 200$ data points,
but this time rather than drawing $y_n \sim
\mbox{bernoulli}(\mbox{logit}^{-1}(\alpha + x_n \, \beta),$ we will
add noise to our estimates of $\phi_n$,^[If we do not add noise, the
values of $\alpha$ and $\beta$ will be exactly identified if $N > K +
1$ and the predictors are not perfectly correlated.]
$$
\phi_n = \alpha + x_n \, \beta + \epsilon_n
$$
where
$$
\epsilon_n \sim \mbox{normal}(0, 0.25).
$$
which is roughly give or take 10% for a value of 50% when converted
back to the probability scale; a noise scale of 0.5 rather than 0.25
would lead to an error of plus or minus 20% or so.

Here are the results of estimating with these approaches compared to
the true values.

```{r results = 'hide'}
K <- 50
rho <- 0.9
Sigma <- matrix(0, K, K)
for (i in 1:K) {
  for (j in 1:K) {
    Sigma[i, j] <- rho^abs(i - j)
  }
}
N <- 200
alpha <- rnorm(1, 0, 0.25)
beta <- rnorm(K, 0, 0.25)

N <- 200
N_test <- 500
x <- mvrnorm(N, rep(0, K), Sigma)

y <- rbinom(N, 1, inv_logit(alpha + x %*% beta))
phi <- alpha + x %*% beta + rnorm(N, 0, 0.25)

weighted_model <- stan_model("weighted-logistic-regression.stan")
weighted_fit <-
  sampling(weighted_model, iter = 4000,
           data = list(N = N, K = K, x = x, phi = as.vector(phi)),
           control = list(adapt_delta = 0.99, stepsize = 0.01),
           refresh = 0, open_progress = FALSE)
weighted_draws <- extract(weighted_fit)

unweighted_model <- stan_model("simple-logistic-regression.stan")
sampled_fit <-
  sampling(unweighted_model, iter = 4000,
           data = list(N = N, K = K, x = x, y = y),
           control = list(adapt_delta = 0.99, stepsize = 0.01),
           refresh = 0, open_progress = FALSE)
sampled_draws <- extract(sampled_fit)
```

```{r}
printf("%10s: alpha = %5.2f  beta[1] = %5.2f  beta[2] = %5.2f",
       "true", alpha, beta[1], beta[2])
printf("%10s: alpha = %5.2f  beta[1] = %5.2f  beta[2] = %5.2f",
       "weighted", mean(weighted_draws$alpha),
       mean(weighted_draws$beta[, 1]), mean(weighted_draws$beta[, 2]))
printf("%10s: alpha = %5.2f  beta[1] = %5.2f  beta[2] = %5.2f",
       "sampled", mean(sampled_draws$alpha),
       mean(sampled_draws$beta[, 1]), mean(sampled_draws$beta[, 2]))
```
Even with that much noise added to the log odds in the weighted case,
the estimates from a training set of size 200 are much better, as can
be seen from the handful of coefficients reported above.

The moral of this story is that binary data is very weak, but
probabilistic data is much stronger.


## Weighted evaluation with scoring functions

All of our loss functions generalize to compute expected loss when
faced with data of the form $\phi_n = \mbox{Pr}[y_n = 1 \mid x_n,
\alpha, \beta]$.  For example, given the linear estimate $\hat{y}_n =
\alpha + x_n \, \beta$ of the log odds that $y_n = 1$, all of our loss
functions are minimized when $\phi_n = \mbox{logit}^{-1}(\alpha + x_n
\, \beta).$

This transition from evaluation on a single outcome to weighted
evaluations where the truth has a probability is commonly used in
evaluating forecasts. The idea is that there's a true uncertainty of
$y_n$ given the predictors $x_n$, and we want to evaluate whether our
inference system is estimating this uncertain properly. Lifting the
scoring rules to deal with probabilistic test data yields general
scoring functions as are used in the probabilistic forecasting
literature. In the simplest case of binary distributions, the general
scoring function corresponds to estimating the cross-entropy rate from
the true probability distribution $\mbox{bernoulli}(\phi_n)$ to the
estimated one $\mbox{bernoulli}(\hat{y}_n)$. That is, the log loss
rate estimates the cross-entropy rate, which is the cost of coding
variables $y_n$ using estimates $\hat{y}_n$ which given that log loss
is a proper scoring rule, is minimized when $\hat{y}_n = \phi_n$.

# Conclusion

When the model is well specified for the problem and there is not a
huge amount of data, as in our logistic regression example, it's much
more accurate to use Bayesian inference. With large amounts of data
relative to the number of predictors, approximate methods perform
almost as well as full Bayes.

One may argue that "big data" abound in real applications and "little
data" should not be a problem.  In reality, much of the big data we
have consists of lots of little data.  We may read millions of base
pairs, but the amount of data we might have about two splice variants
of interest is typically on the scale of the simulations done in this
paper.  Similarly, if we look at e-commerce, we have tons of data, but
relatively little data about individual products or customers, as
new users and items are continually added and old behaviors change.

To summarize the quantitative results in this short note, the error
rates for a training set of size `r N` with `r K` correlated
predictors are summarized in the following table.

| Inference            | root mean square error            | log loss rate                |
|---------------------:|----------------------------------:|-----------------------------:|
| Full Bayes           | `r sqrt(sq_error_bayes / N_test)` | `r log_loss_bayes / N_test` |
| MAP (posterior mode) | `r sqrt(sq_error_map / N_test)`   | `r log_loss_map / N_test`      |
| VB (posterior mean)  | `r sqrt(sq_error_vi / N_test)`    | `r log_loss_vi / N_test`     |

The result from taking a single posterior draw $(\alpha^{(m)},
\beta^{(m)})$ as a point estimate depends heavily on the draw, but
almost all such draws are worse than the posterior mode and the
posterior mean.


## Acknowledgements {-}

Thanks to the Mark Johnson for inspiring to write up an example of
where full Bayesian inference was better than point estimates.  Thanks
also to all the commenters in our Bayesian computation reading group.
Thanks in particular to Yuling Yao for all of the clarifying
questions;  I tried to answer them all here.


# Appendix A:  Some alternative evals

We also include a table of evaluation results for full Bayes versus
plugging in either the posterior mode (MAP) or mean (VB).  We only run
a single simulation for each set of conditions as it provides an
overall feel for the trend of the evaluations.^[Combining simulations
is tricky as each simulated condition has a different underlying
entropy and hence different lower bound on cross-entropy and hence log
loss.]


```{r}
cov_matrix <- function(K, rho) {
  Sigma <- matrix(0, K, K)
  for (i in 1:K) {
    for (j in 1:K) {
      Sigma[i, j] <- rho^abs(i - j)
    }
  }
  Sigma
}

losses <- function(y, y_hat) {
  N <- length(y)

  abs_err <- sum(abs(y - y_hat))
  mae <- abs_err / N

  sq_err <- sum((y - y_hat)^2)
  rmse <- sqrt(sq_err / N)

  log_loss <- -sum(dbinom(y, 1, y_hat, log = TRUE))
  log_loss_rate <- log_loss / N

  list(mae = mae, rmse = rmse, log_loss_rate = log_loss_rate)
}

losses_table <- function(y, y_hat, N, K, rho, estimator, rep_in) {
  losses = losses(y, y_hat)
  data.frame(estimator = estimator, N = N, K = K, rho = rho,
             log_loss_rate = losses$log_loss_rate,
             rmse = losses$rmse,
             mae = losses$mae,
	     rep = rep_in)
}

model <- stan_model("logistic-regression.stan")

Ns <- c(8, 16, 32, 64, 128, 256, 512) # , 32, 64, 128, 256, 512, 1024)
Ks <- c(2, 8, 32, 64, 256)
rhos <- c(0)  # , 0.9)

results <- data.frame()

for (N in Ns) {
  for (K in Ks) {
    for (rho in rhos) {
      for (rep in 1:10) {
        cat("N = ", N, ";  K = ", K, ";  rho = ", rho, "; rep = ", rep, "\n")
        Sigma <- cov_matrix(K, rho)

        alpha <- rnorm(1, 0, 0.5)
        beta <- rnorm(K, 0, 0.5)

        x <- mvrnorm(N, rep(0, K), Sigma)

        y <- rbinom(N, 1, inv_logit(alpha + x %*% beta))
        N_test <- 100
        x_test <- mvrnorm(N_test, rep(0, K), Sigma)
        y_test <- rbinom(N_test, 1, inv_logit(alpha + x_test %*% beta))

        data <- list(K = K, N = N, x = x, y = y,
                     N_test = N_test, x_test = x_test, y_test = y_test)

        # Bayes
        fit <- sampling(model, data = data, refresh = 0, init = 0.5,
                        chains = 4, iter = 4000, control = list(stepsize = 0.01, adapt_delta = 0.95),
                        open_progress = FALSE)

        y_test_hat_bayes <- rep(0, N_test)
        for (n in 1:N_test) {
          y_test_hat_bayes[n] <- mean(extract(fit)$E_y_test[ , n])
        }
	bayes_table <- losses_table(y_test, y_test_hat_bayes, N, K, rho, "Bayes", rep)


        # MAP plug-in
        fit_map <- optimizing(model, data = data, refresh = 0, verbose = FALSE)
        alpha_star <- fit_map$par[['alpha']]
        beta_star <- rep(0, K)
        for (k in 1:K)
          beta_star[k] <- fit_map$par[[paste('beta[', k, ']', sep = "")]]

        y_test_star <- inv_logit(alpha_star + (x_test %*% beta_star))
        map_table <- losses_table(y_test, y_test_star, N, K, rho, "MAP", rep)

        # VB plug-in
        alpha_hat <- mean(extract(fit)$alpha)
        beta_hat <- rep(0, K)
        for (k in 1:K)
          beta_hat[k] <- mean(extract(fit)$beta[ , k])
        y_test_hat_vb <- inv_logit(alpha_hat + (x_test %*% beta_hat))
	vb_table <- losses_table(y_test, y_test_hat_vb, N, K, rho, "VB", rep)

        # reduce to difference in error from Bayes to standardize
        map_table$log_loss_rate <- map_table$log_loss_rate - bayes_table$log_loss_rate
	map_table$rmse <- map_table$rmse - bayes_table$rmse
	map_table$mae <- map_table$mae - bayes_table$mae

        vb_table$log_loss_rate <- vb_table$log_loss_rate - bayes_table$log_loss_rate
	vb_table$rmse <- vb_table$rmse - bayes_table$rmse
	vb_table$mae <- vb_table$mae - bayes_table$mae

        bayes_table$log_loss_rate <- bayes_table$log_loss_rate - bayes_table$log_loss_rate
	bayes_table$rmse <- bayes_table$rmse - bayes_table$rmse
	bayes_table$mae <- bayes_table$mae - bayes_table$mae

        results <- rbind(results, map_table)
        results <- rbind(results, vb_table)
        results <- rbind(results, bayes_table)
      }	
    }
  }
}

mae_results <- aggregate(mae ~ estimator + N + K + rho, results, mean)
log_loss_rate_results <- aggregate(log_loss_rate ~ estimator + N + K + rho, results, mean)
rmse_results <- aggregate(rmse ~ estimator + N + K + rho, results, mean)

mae_results <- cbind(mae_results,
                     loss_type = rep("mae", dim(mae_results)[1]))
log_loss_rate_results <- cbind(log_loss_rate_results,
                               loss_type = rep("log loss", dim(log_loss_rate_results)[1]))
rmse_results <- cbind(rmse_results,
                      loss_type = rep("rmse", dim(mae_results)[1]))

names(mae_results)[5] <- "loss"
names(log_loss_rate_results)[5] <- "loss"
names(rmse_results)[5] <- "loss"

plot_df <- rbind(log_loss_rate_results, rmse_results)  # , mae_results
sub_df <- subset(plot_df, rho == 0)
sub_df$rho <- NULL

results_plot <-
  ggplot(sub_df,
         aes(x = N, y = loss, color = estimator, group = estimator)) +
  geom_line(aes(group = estimator, color = estimator)) +
  scale_x_log10(breaks = Ns) + 
  facet_grid(loss_type ~ K, scales = "free_y") +
  ylab("loss - Bayes loss") +
  ggtheme_tufte()
results_plot
```

# Appendix B:  R Session Information

```{r}
sessionInfo()
```